---
title: "Cyclistic_bike_share_analysis"
author: "Sudharshan"
date: "`r Sys.Date()`"
output: html_document
---

## The ask phase

**The ask phase of the case study focuses on**

  * The problem statement
  * Effect of your insight on business decision
  * Business task
  * Stakeholders

**Business task**

Conduct data analysis to identify the difference between how casual riders and annual members use the bike

**Stakeholder analysis**

  * Responsible: Sudharsana Prasad
  * Consult: Marketing team
  * Accountable: Lily Moreno
  * Inform: Executives

## Prepare

The prepare phase of the case study focuses on:
  * Data source information
    ** Location
    ** How it is organized
    ** ROCCC, integrity and biases
    ** Licensing, privacy, accessibility and security
    ** Relevance to business task
    ** Problems associated with the data  

**Date location**

The data is stored at [Index of bucket](https://divvy-tripdata.s3.amazonaws.com/index.html). The data we are interested in is from *from: 2023/01 to: 2023/12*. The data is avilable in zip format and we have extracted all the required datasets to [trip_data_2024](file:///D:/DA%20tutorial/Case%20studies/Cyclistic%20data%20sources/trip_data_2024/)

**Data cleaning and inspection**

The first step is to read all the datasets into R

```{r Loading packages}
#install.packages("tidyverse", repos = "http://cran.us.r-project.org")
#install.packages("skimr", repos = "http://cran.us.r-project.org")
#install.packages("DescTools", repos = "http://cran.us.r-project.org")
#install.packages("dplyr", repos = "http://cran.us.r-project.org")
#install.packages("ggplot2", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(skimr)
library(DescTools)
library(dplyr)
library(ggplot2)
```

```{r Loading datasets into R}
parent_dir <- "D:\\DA tutorial\\Case studies\\Cyclistic data sources\\trip_data_2023\\"
for(i in 202301:202312){
  file_name <- paste0(parent_dir, i, "-divvy-tripdata\\", i, "-divvy-tripdata.csv")
  assign(paste0("trip_", i), read_csv(file_name))
}
```
**Inspecting data**
The trip data is stored in 12 different tables, each table for every month in 2023. The tables are read into 12 different data frames from trip_202301 to trip_202312.  
There are totally 13 columns and three different data types (Character, Double and Date time). The data frames are now combined into a single data frame trips_2023
```{r Combine all data frames}
trips_2023 <- rbind(trip_202301, trip_202302, trip_202303, trip_202304, trip_202305, trip_202306, trip_202307, trip_202308, trip_202309, trip_202310, trip_202311, trip_202312)
```

## Process

Calculating how long the rides were for each ride in 2023.
```{r Calculating ride_length}
trips_2023 <- mutate(trips_2023, ride_length = trips_2023$ended_at - trips_2023$started_at)
str(trips_2023)
```
Creating a column of day_of_week and removing incomplete observations
```{r Create a column for day of the week}
trips_2023 <- mutate(trips_2023, day_of_week = weekdays(trips_2023$started_at))
str(trips_2023)
trips_2023 <- na.omit(trips_2023)
```
Arranging the data frame in ascending order of started_at column
```{r Arrage data frame in chronological order}
trips_2023 <- arrange(trips_2023, started_at)
```
## Analyze

Calculating mean, and max of ride_length
```{r Mean and max of ride_length}
mean(trips_2023$ride_length)
max(trips_2023$ride_length)
```
Calculating the Mode of day_of_week
```{r Calculating Mode of Day of the week}
Mode(trips_2023$day_of_week) # Package used DescTools
```

### Calculations

Calculate the average ride_length for members and casual riders for each month
```{r Monthly average ride length calculation}
trips_2023 %>% 
  group_by(month(started_at), member_casual) %>% 
  summarise(mean(ride_length))
```
Calculate the average ride_length for users by day_of_week.
```{r Avergae ride lenght by day of the week}
trips_2023 %>% 
  group_by(month(started_at), member_casual, day_of_week) %>% 
  summarize(mean(ride_length))
```
Calculate the number of rides for users by day_of_week
```{r  Numer of rides by day of the week}
rides_table <- trips_2023 %>% 
  group_by(month(started_at), member_casual) %>%
  count(member_casual)
colnames(rides_table) <- c('month', 'member_type', 'num_of_rides')
tibble(rides_table)
```
## Share

### Visualization
```{r Num of casual and member riders in 2023}
ggplot(trips_2023) +
  geom_bar(mapping = aes(x = member_casual, fill = day_of_week)) +
  labs(title = "Casual and memeber riders count in 2023", subtitle = "Day of the week", x = "Type of rider", y = "Number of rides")
```

```{r Trend of causual and memeber riders in 2023}
trips_2023 %>% ggplot(aes(x = fct_relevel(as.character(month(started_at), c("10", "11", "12"), after = 12)))) +
  geom_bar() +
  facet_wrap(~member_casual) +
  labs(title = "Monthly rides", x = "Month", y = "number of rides" )
```


## Findings

  * Casual rides use the bikes on weekends, while members use the bike mostly on weekdays
  * The number of rides go up during May, June and July. The peak is in July and drops during the winter months
  * The number of Members is more than the number of casual riders.
